---
title: Practical Machine Learning Project
author: "Nicolas"
output:
  html_document:
    fig_height: 10
    fig_width: 10
---

## Get and Clean Data  
In this project, we are using data from the accelerometers on the belt, forearm, arm, and dumbell of six participants to predict how well they did the exercise.  
```{r, cache = T}
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)
```
### Download the Data
```{r, }
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainCSV <- "./pml-training.csv"
testCSV  <- "./pml-testing.csv"
if (!file.exists(trainCSV)) {
  download.file(trainUrl, destfile=trainCSV, method="curl")
}
if (!file.exists(testCSV)) {
  download.file(testUrl, destfile=testCSV, method="curl")
}
```  
### Read the Data
```{r, cache = T}
trainRaw <- read.csv("./pml-training.csv")
testRaw <- read.csv("./pml-testing.csv")
dim(trainRaw)
dim(testRaw)
```
The training dataset contains 19622 observations and 160 variables, while the testing dataset contains 20 observations and 160 variables. The "classe" variable in the training dataset is the outcome to predict. 
### Clean the data
Remove observations with missing values and non useful variables.
```{r, cache = T}
sum(complete.cases(trainRaw))
```
Remove columns that contain NA missing values.
```{r, cache = T}
trainRaw <- trainRaw[, colSums(is.na(trainRaw)) == 0] 
testRaw <- testRaw[, colSums(is.na(testRaw)) == 0] 
```  
Remove columns that do not contribute to the accelerometer measurements.
```{r, cache = T}
classe <- trainRaw$classe
trainRemove <- grepl("^X|timestamp|window", names(trainRaw))
trainRaw <- trainRaw[, !trainRemove]
trainCleaned <- trainRaw[, sapply(trainRaw, is.numeric)]
trainCleaned$classe <- classe
testRemove <- grepl("^X|timestamp|window", names(testRaw))
testRaw <- testRaw[, !testRemove]
testCleaned <- testRaw[, sapply(testRaw, is.numeric)]
```
The clean training dataset contains 19622 observations and 53 variables. 
The testing dataset contains 20 observations and 53 variables. 
### Split the training data
Let's split the clean training dataset into training dataset at 70% and a validation dataset at 30%. 
The validation dataset will be used to conduct cross validation before the testing dataset.  
```{r, cache = T}
set.seed(22519) # For reproducibile purpose
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainData <- trainCleaned[inTrain, ]
valData <- trainCleaned[-inTrain, ]
```

## Data Modeling
Fit a predictive model for the activity recognition using **Random Forest** algorithm as it automatically selects important variables and it is robust to correlated covariates and outliers. 
```{r, cache = T}
controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf
```
Estimate the performance of the model on the validation dataset.  
```{r, cache = T}
predictRf <- predict(modelRf, valData)
confusionMatrix(valData$classe, predictRf)
```
```{r, cache = T}
accuracy <- postResample(predictRf, valData$classe)
accuracy
err <- 1 - as.numeric(confusionMatrix(valData$classe, predictRf)$overall[1])
err
```
The estimated accuracy of the model is 99.42% and the estimated out of sample error is 0.58%.

## Predicting on Test Dataset
Apply the model to the original testing dataset.  
```{r, cache = T}
result <- predict(modelRf, testCleaned[, -length(names(testCleaned))])
result
```  

## Figures:
Correlation Matrix Visualization  
```{r, cache = T}
corrPlot <- cor(trainData[, -length(names(trainData))])
corrplot(corrPlot, method="color")
```
Decision Tree Visualization
```{r, cache = T}
treeModel <- rpart(classe ~ ., data=trainData, method="class")
prp(treeModel) # fast plot
```